Stability of variational quantum dynamics with Gaussian approximants

### 2013-11-19 paper.tex
writing

\def\G{|Γ〉}
\def\Gx{|Γ_χ〉}
\def\cite#1{$[\hbox{\tt #1}]$}
\def\T{^{\rm T}}

@@

When they realised that positions and momenta are essentially uncertain, the first quantum physicists were bound to discover the gaussian wave packet by seeking conditions that produce a normal distribution of probability.  They soon understood the free particle and turned to the harmonic oscillator, investigating the form of its ground state and the way its excited states generate sinusoidal motion in the classical limit.  These problems were also solved by the gaussian wave packet.  With its width fixed by equipartition of energy and the uncertainty principle, it became the coherent state.

The simplest mechanical systems allow Schrödinger's law of motion to be solved in closed form, but most quantum problems must be solved approximately.  This can be done variationally, specifying a class of approximants that can be manipulated exactly, then finding one of them that most nearly solves the problem.  To measure how well a wave function $|ψ(t)〉$ approximates the dynamics of a system, we will compare the left and right sides of Schrödinger's equation, $i\hbar{d|ψ〉\over dt}$ and $H|ψ〉$; these are both vectors in Hilbert space, and the norm defines a distance between them.  It remains to choose a class of approximants.  The gaussian wave packets are an common choice, because they differ in their scale and their central position and momentum, useful properties to determine for any quantum state.

The invention of the laser in 1963 called for a quantum theory of intense radiation.  Laser beams are hard to analyse because their quantum descriptions are so large, comprising up to $10^{18}$ photons distributed across thousands of modes.  The theory was made managable by Glauber and Sudarshan, who showed that any quantum state can be expanded as a superposition of coherent states.  Laser light is very coherent, so these expansions often represent it efficiently.  Moyal(?) developed an efficient way to calculate dynamics, leading to a family of phase space Monte-Carlo methods.  When very coherent matter, in the form of atomic Bose-Einstein condensates, became easy to access in the lab, phase space methods were sucessfully applied to calculate its dynamics, and have since been applied to other forms of condensed matter.  They have solved many problems, but are limited by intractable convergence issues.

In the 1970s, Heller put the two ideas together, proposing to simulate quantum dynamics variationally, using finite superpositions of coherent states as the approximants.  This is an appealing approach.  Every quantum state can be approximated arbitrarily well as the number of components in the superposition increases.  The experience of quantum optics shows that, in many systems, the approximation converges with a reasonable number of components.  Also, variational dynamics should avoid the covergence problems that limit the Monte-Carlo methods.

@@

Since Heller suggested it, there have been several attempts to apply this numerically, but none has entirely succeeded.  In all cases, numerical instability has prevented the superposition being expanded to the point of convergence.
%% Consequence
These attempts have aimed to solve chemistry problems, involving dozens of atoms, where the lack of convergence limited the precision that could be obtained, but superpositions with a few components, and ad-hoc methods to improve the stability, still gave useful results.  The kind of problems that people approach with phase space methods come from atom optics or condensed matter; they involve thousands or millions of atoms, and the convergence problems would prevent any useful information being obtained.  A systematic approach to regularisation is required in order to approach these problems variationally.  The prospective rewards are great: these problems include high temperature superconductivity and the dynamics of the early universe, topics on which today's physicists expend massive efforts.

@@

Both condition and numerical stability refer to reasons that a computer can give the wrong answer to a numerical problem.  The distinction between them is a bit subtle.  A problem is ill-conditioned if the solution is very sensitive to the data with which the problem is stated: the canonical example is computing $\sin(x)$ for $x=10⁶$, whence a relative error of $π×10^{-6}$ in $x$ causes a 100\% error in $\sin(x)$.  The key thing to note is that this has nothing to do with how $\sin(x)$ is computed—the error in $x$ could be due to measurement uncertainty, in which case there would be no way to calculate $\sin(x)$: we'd need to measure it directly.  Condition is a property of problems.  Stability is a property of solutions.  For example, if we asked a computer to calculate $1-\cos(x)$ in the obvious way, with $x$ between 0 and $10^{-10}$, the result might vary randomly around 0, in a range determined by the precision with which the computer stored numbers.  But this problem is not intrinsically hard: at any reasonable level of precision, it amounts to finding $x²/2$.  The method of subtracting a numerical value of $\cos(x)$ from 1, which generates random numbers, is unstable.

The instability appears to have two major causes.  The first problem occurs when quantum states are expanded over a finite set of coherent states.  There is a natural way to expand a state as an integral over the coherent states, with a weight function that varies smoothly with the coherent amplitude.  When this function is sampled at a discrete set of coherent amplitudes and the integral approximated by a sum, the state is approximated to some precision.  Part of the error in this appoximation is due to the finite set of coherent states having bounded amplitude, wheras the state being approximated might have components with arbitrarily large particle number—for instance, it might be a coherent state.  When expansion weights for the finite set of coherent states are set by least squares, components with large particle number are approximated by adjacent amplitudes having large weights, with opposite signs so that components with low particle numbers cancel.  This causes the expansion weights to be large, to not resemble samples of any smooth function, and to change sharply as the state changes.  All these properties make it difficult to numerically integrate a rate of change that has been expanded by least squares.  The problems are exacerbated when the amplitudes are spaced irregularly, instead of on a grid.

@@

\beginsection{Gabor frames}

>>>>

\def\G{|Γ〉}
\def\A{|A〉}
\def\Gx{|Γ_χ〉}
\def\cite#1{$[\hbox{\tt #1}]$}
\def\T{^{\rm T}}

@@

The Gaussian wave packet is a ubiquitous idea in quantum mechanics.  Its discovery was inevitable: people were provoked by understanding that observables are essentially uncertain to study physical states in which they are normally distributed.  The Gaussian wavepacket also crops up as the quantum state which, placed in a quadratic potential, best approximates the harmonic oscillation of classical mechanics.  In this potential, there is a natural width for the packet, which ensures its shape remains constant during its motion.  This width can also be set by requiring equipartition of energy and the uncertainty principle.  Wavepackets with this natural width are coherent states.

It has also been a productive tool.  The simplest mechanical systems allow Schrödinger's law of motion to be solved in closed form, but most quantum problems must be solved approximately.  This can be done variationally, specifying a class of approximants that can be manipulated exactly, then finding one of them that most nearly solves the problem.  If the problem is to find a wave function $|ψ(t)〉$ that evolves according to a Hamiltonian $H$, the quality of approximation can be measured by comparing the left and right sides of Schrödinger's equation, $i\hbar{d|ψ〉\over dt}$ and $H|ψ〉$.  These are both vectors in Hilbert space, and can be compared by the norm of their difference.  The gaussian wave packets have often been used as variational approximants, because they are defined by their width and their central position and momentum, useful properties to determine for any quantum state.

The invention of the laser in 1963 called for a quantum theory of intense radiation.  Laser beams are hard to analyse because their quantum descriptions are so large, comprising up to $10^{18}$ photons distributed across thousands of modes.  The theory was made managable by Glauber and Sudarshan, who showed that any quantum state can be expanded as a superposition of coherent states.  Laser light is very coherent, so these expansions often represent it efficiently.  Moyal(?) developed an efficient way to calculate dynamics, leading to a family of phase space Monte-Carlo methods.  When very coherent matter, in the form of atomic Bose-Einstein condensates, became easy to access in the lab, phase space methods were sucessfully applied to calculate its dynamics.  They have since been applied to other forms of condensed matter.  They have solved many problems, but are limited by intractable convergence issues.

In the 1970s, Heller put variational mechanics and phase space expansions together.  He proposed to simulate quantum dynamics variationally, using finite superpositions of coherent states as the approximants.  This is an appealing approach.  Every quantum state can be approximated arbitrarily well as the number of components in the superposition increases, and the finite superposition converges to a continuous phase space expansion.  The experience of quantum optics shows that, in many systems, the expansion converges when the superposition has a reasonable number of terms.  Also, variational dynamics should avoid the covergence problems that limit the Monte-Carlo methods.

@@

There have been several attempts to implement Heller's proposal, but none has entirely succeeded.  In all cases, numerical instability has prevented the superposition being expanded to the point of convergence.
%% Consequence
These attempts have aimed to solve chemistry problems, involving dozens of atoms.  Their failure to converge limited their precision, but useful results were obtained from superpositions with a few components, and ad-hoc methods to improve their stability.  The problems that people attempt with phase space methods come from atom optics or condensed matter, and involve thousands or millions of atoms.  In these systems, the instability would prevent any useful information being obtained.  Therefore a systematic approach to regularisation is required in order to approach these problems variationally.  The prospective rewards are significant: many-particle problems include high temperature superconductivity and the dynamics of the early universe.

@@

Both condition and numerical stability refer to reasons that a computer can give the wrong answer to a numerical problem, and the distinction between them is a bit subtle.  A problem is ill-conditioned if the solution is very sensitive to the data with which the problem is stated: the canonical example is computing $\sin(x)$ for $x=10⁶$, whence a relative error of $π×10^{-6}$ in $x$ causes a 100\% error in $\sin(x)$.  The key thing to note is that this has nothing to do with how $\sin(x)$ is computed—the error in $x$ could be due to measurement uncertainty, in which case there would be no way to calculate $\sin(x)$: we'd need to measure it directly.  Condition is a property of problems.  Stability is a property of solutions.  For example, if we asked a computer to calculate $1-\cos(x)$ in the obvious way, with $x$ between 0 and $10^{-10}$, the result might vary randomly around 0, in a range determined by the precision with which the computer stored numbers.  But this problem is not intrinsically hard: at any reasonable level of precision, it amounts to finding $x²/2$.  The method of subtracting a numerical value of $\cos(x)$ from 1, which generates random numbers, is unstable.  Experiments with many-particle systems suggest that phenomena such as interference and superconductivity are robust, and changing the system slightly does not greatly change the phenomena that result. So computing the properties of these systems should be a well-conditioned problem.  However, methods of solving them variationally, using coherent states, have been very unstable.

The instability appears to have two major causes.  The first problem occurs when quantum states are expanded over a finite set of coherent states.  There is a natural way to expand a state as an integral over the phase space of coherent states, with a weight function that varies smoothly with the coherent amplitude.  When this function is sampled at a discrete set of coherent amplitudes and the integral approximated by a sum, the state is approximated to some precision.  Part of the error in this appoximation is due to the finite set of coherent states having bounded amplitude, wheras the state being approximated might have components with arbitrarily large particle number.  When expansion weights for the finite set of coherent states are set by least squares, components with large particle number are approximated by adjacent amplitudes having large weights, with opposite signs so that components with low particle numbers cancel.  This causes the expansion weights to be large, to not resemble samples of any smooth function, and to change sharply as the state changes.  All these properties make it difficult to numerically integrate a rate of change that has been expanded by least squares.  The problems are exacerbated when the amplitudes are spaced irregularly, instead of on a grid.

@@

\beginsection{Frames}

Because we are after numerical methods, our goal is for expansions such as $\G f$ and $\A d$ to be numerically stable.  This means, for one thing, that a small change in $f$ should cause a small change in $\G f$.  The coherent states that compose $\G$ are overcomplete, so fixing the value of $\G f$ does not fix the value of $f$.  However, in order to compute with this expansion, we need to find a way of solving for $f$ such that a small change in $\G f$ causes a small change in the  $f$ that is the result of our method.

There is a precise way to state these conditions.  For the expansion $\G$ to be stable, its columns must form a frame\cite{1992-Daubechies-Ten}.  This means, in the continuous case, that there are constants $A$ and $B$ such that
$$ A‖|ψ〉‖²≤\int |〈γ|ψ〉|²\,dγ≤B‖|ψ〉‖²,$$
for all kets $|ψ〉$.  In the discrete case, the integral is replaced by a sum.  Because the operator $\G$ is linear, the ket $|ψ〉$ could be a difference $|φ₁〉-|φ₂〉$.  In this case, the right hand inequality guarantees that a small change in a ket will produce a small change in the components $〈γ|ψ〉$, which the left hand inequality says tht a small change in the components will produce a small change in the expanded ket.

In the particular case of the coherent states, we have
$$ \int |〈γ|ψ〉|²\,dγ=〈ψ|\int |γ〉〈γ|\,d²γ\,|ψ〉=π,$$
so the frame condition is satisfied, with $A=B=π$.  A set of this kind, where the upper and lower frame bounds are the same, is known as a tight frame.  Mathematicians have known this for a long time, and refer to the coherent states, and other frames that result from translating and modulating functions that aren't necessarily Gaussian, as Gabor frames.

(Talk about dual frames, minimal norm expansions.)

In the numerical case, the number of frame kets is finite.  Unless $\A$ is singular, the frame bounds always exist, with $A$ and $B$ something like the largest and smallest singular values of $\A$.  So we need to play a bit loose with the mathematical definitions.  For numerical purposes, a bound of the order $10^{15}$ might as well be infinite, and one of the order $10^{-15}$ might as well be zero.  Also, a finite set of states can not  be a frame for the whole of Hilbert space: there will always be directions orthogonal to all the components, for which $∑_r|〈α_r|ψ〉|²=0$.  So we need to think carefully about what space $|ψ〉$ can lie in.  The obvious choice would be the range of $\A$, the span of the $|α_r〉$.

But this causes problems with the lower frame bound.  The amplitudes of the coherent states will be bounded: suppose $|α_r|≤a$.  A finite set of coherent states with different amplitudes are linearly independent, so the simultaneous equations $〈n|A〉d=0$, where n ranges from $0$ to $R-1$, have a solution.  This means that $\A d$ is a state of the form $c_R|R〉+c_{R+1}|R+1〉+⋯$.  Now, the component of $|n〉$ in $|a_r〉$ has modulus no greater than $a^n/\sqrt{n!}$, so $|c_n| ≤ a^n/\sqrt{n!} ‖c‖₁$.  This means that, in the limit of dense amplitudes, where $R$ increases while $a$ stays constant, we can find vectors $d$ such that $‖\A d‖$ tends to zero.  (Fix that to argue the right way round.)  There are states in the span of $\A$ with large particle number, that barely overlap any of the states $|α_r〉$, and have very small components along the frame directions.  So, from a numerical point of view, dense, bounded sets of coherent states fail to be frames.  If the $α_r$ extended over the whole of phase space, the set would include some larger amplitudes for which $〈α_r|ψ〉$ was not small.

The obvious solution is to not attempt to expand the entire span of $\A$, but first project onto some subspace for which the $|α_r〉$ are a frame with numerically reasonable bounds.  This is what regularisation methods do.  In particular, singular vectors of $\A$ with large singular values span a space for which the states are a frame.  In practice, these correspond to states with limited particle number.  So this might be why projecting onto smal l particle number makes simulations more stable: the ensemble of coherent states is a frame for that space!  It also suggest that, for accurate and stable expansions, we want coherent amplitudes large enough that the components don't have a great overlap with the state being expanded.

It is interesting to consider the squeezed states.  Mathematicians have thought about two types of frames: Gabor frames, generated by translation and momentum kicks, and wavelet frames, generated by translation and limited types of squeezing.  It is unclear whether anyone has studied if the squeezed states are a frame.  If they are not, it might explain why numerical expansions over squeezed states have proven to be so unstable.
